Fast Sorting on Modern Computers

### 2005-08-01 docs/tech_report/merge.tex
minor edits to fix latex errors
added testing comments

\section{Results}
\subsection{Test Parameters}
The level 1 cache is 8k, so 1024 keys from each array can be sorted in the level
1 cache. 262144 keys can be sorted in the level 2 cache. When arrays were
pre-sorted, they were pre-sorted to a size four, or eight if the number of
passes was going to be odd. These were reversed for the double-aligned
multi-mergesort; since it does a k-way merge into the source array, the sorted
cache sized segments need to be in the auxiliary array. An 8-heap was used for
the multi-mergesorts.

@@

Despite the improvement in cache misses due to multi-mergesort, it still cannot
compensate for the increase in instruction count as a result, and it has a higher
cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
other sorts generally perform in $O(NlogN)$ time, with double-aligned tiled mergesort
being the fastest. Note that the alignment of the mergesort may not work as
desired due to the size of the cache. This was optimized for a 2MB cache, while
the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to
the same cache block. However, since the cache is 8-way associative, this is not
serious, and should not result in a lot of thrashing.

@@

Overall, when the k-ary heap is used, there are the same number of branch
misses. This means that using the technique to reduce cache misses does not
adversely affect branch prediction rates (it does, however, still affect the
instruction count).  The rate of branch misses is much the same as that in
heapsort, and progresses in the same manner (see Figure \ref{memory-tuned heapsort (8-heap)
curve}). Note that \n{comparison 0} occurs every time, and would be optimized
out by the compiler.

>>>>

\section{Results}
\subsection{Test Parameters}
%TESTING
The level 1 cache is 8k, so 1024 keys from each array can be sorted in the level
1 cache. 262144 keys can be sorted in the level 2 cache. When arrays were
pre-sorted, they were pre-sorted to a size four, or eight if the number of
passes was going to be odd. These were reversed for the double-aligned
multi-mergesort; since it does a k-way merge into the source array, the sorted
cache sized segments need to be in the auxiliary array. An 8-heap was used for
the multi-mergesorts.

@@

Despite the improvement in cache misses due to multi-mergesort, it still cannot
compensate for the increase in instruction count as a result, and it has a higher
cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
other sorts generally perform in $O(NlogN)$ time, with double-aligned tiled mergesort
being the fastest. Note that the alignment of the mergesort may not work as
%TESTING
desired due to the size of the cache. This was optimized for a 2MB cache, while
the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to
the same cache block. However, since the cache is 8-way associative, this is not
serious, and should not result in a lot of thrashing.

@@

Overall, when the k-ary heap is used, there are the same number of branch
misses. This means that using the technique to reduce cache misses does not
adversely affect branch prediction rates (it does, however, still affect the
instruction count).  The rate of branch misses is much the same as that in
heapsort, and progresses in the same manner (see Figure \ref{memory-tuned
heapsort curve}). Note that \n{comparison 0} occurs every time, and would be
optimized out by the compiler.

### 2005-08-01 docs/tech_report/method.tex
more minor renaming
added testing comments

The SimpleScalar settings were chosen to be similar to LaMarca's simulations.
LaMarca used the \n{Atom} simulator, which was used to test software for the DEC
AlphaStation. It simulated a 2MB cache with a 32 byte cache line. The remainder
of the settings were taken from the DEC Alpha: 8KB level 1 data cache, 8KB level
1 instruction cache and shared instruction and data level 2 cache.

\subsection{Gnuplot}
\label{Gnuplot}
\cc{Gnuplot} is a tool for plotting data files. It provides visualisations of data in
two and three dimensions, and is used to create all of the graphs in this
document. Its homepage is \n{http://www.gnuplot.info/}.

@@

The main program provided the framework for the development of the sorts.
Imaginatively called \cc{main.c}, this program provided the means for testing
and (roughly) timing each sort. Figure \vref{Key sections of Main.c} contains
this code.

@@

\subsection{Valgrind}
\label{Valgrind}
\cc{Valgrind} is a set of tools which are used for debugging and profiling.  By
default it works as a memory checker, and detects errors such as the use of
uninitialised memory and accessing memory off the end of an array. This use
makes it particularly useful for bounds checking. No configuration was required
for this, and simply running \cc{valgrind a.out} ran the program through
valgrind. The homepage for valgrind is at \n{http://valgrind.kde.org/}.

\subsection{PapiEx}
\cc{PapiEx} is a tool used to access performance counters on a Pentium 4. It
uses the PAPI framework to count certain characteristics across a program's
execution. We used it to count the number of cycles a sort took to run. This
provided an accurate measure of how fast a sort executes on a Pentium 4. The
PapiEx homepage is at \n{http://icl.cs.utk.edu/~mucci/papiex/}.

@@

LaMarca's final step was to create a multi-mergesort. This attempted to reduce
the number of cache misses in the final merge steps of tiled mergesort by
merging all the sorted arrays in one step. This required the use of an 8-heap,
which had been developed as part of heapsort.
%TODO this isnt what happened exactly
Once these sorts were written it was attempted to improve them slightly.
Instead of a tiled mergesort, a double-aligned tiled mergesort was written.  This
attempted to completely sort in the level 1 cache, then to completely sort in
the level 2 cache. Double multi-mergesort was a combination of this and of
multi-mergesort. The rewrite was based on the base mergesort, and was written
far more cleanly, being more readable and efficient. These were also found to
reduce the instruction count, as well as the level 2 cache miss count. These
results are discussed in more detail in Section \ref{Mergesort results}.

@@

%TODO another sentence here
Each of the four $O(NlogN)$ sorts, radixsort and shellsort, was treated in this fashion.

@@

%TODO a lot of this may be in the first sorts chapter, maybe heapsort.
%discuss: BIG + random.org, 64 bit in lamarca vs 32 bit, do_nothing and how it
%affects cache hits, scaling in the software branch predictors. How to compare
%and contrast varying results (branch vs softwre branhc; P4 vs simplescalar).
%What factors are used for testing: /proc/cpuinfo output, cahce and branch
%predictor data from P4, simplescalar inputs, software predictor, verfication of
%sorts.

@@

As well as the number of cycles, Pentium 4 performance registers can also provide
most of the same information as SimpleScalar, with different parameters. For
example, the cache line of a Pentium 4 machine is 128 bytes whereas 32 byte
cache lines were used in SimpleScalar. The addition of real world data would
strengthen our results.

>>>>

%TESTING - also what version
The SimpleScalar settings were chosen to be similar to LaMarca's simulations.
LaMarca used the \n{Atom} simulator, which was used to test software for the DEC
AlphaStation. It simulated a 2MB cache with a 32 byte cache line. The remainder
of the settings were taken from the DEC Alpha: 8KB level 1 data cache, 8KB level
1 instruction cache and shared instruction and data level 2 cache.

%TESTING - what versino
\subsection{Gnuplot}
\label{Gnuplot}
\cc{Gnuplot} is a tool for plotting data files. It provides visualisations of data in
two and three dimensions, and is used to create all of the graphs in this
document. Its homepage is \n{http://www.gnuplot.info/}.

@@

The main program provided the framework for the development of the sorts.
Imaginatively called \cc{main.c}, this program provided the means for testing
and (roughly) timing each sort. Figure \vref{Key sections of Main.c} contains a
sample of this code.

@@

%TESTING - what version
\subsection{Valgrind}
\label{Valgrind}
\cc{Valgrind} is a set of tools which are used for debugging and profiling.  By
default it works as a memory checker, and detects errors such as the use of
uninitialised memory and accessing memory off the end of an array. This use
makes it particularly useful for bounds checking. No configuration was required
for this, and simply running \cc{valgrind a.out} ran the program through
valgrind. The homepage for valgrind is at \n{http://valgrind.kde.org/}.

%TESTING - what version
\subsection{PapiEx}
\cc{PapiEx} is a tool used to access performance counters on a Pentium 4. It
uses the PAPI framework to count certain characteristics across a program's
execution. We used it to count the number of cycles a sort took to run. This
provided an accurate measure of how fast a sort executes on a Pentium 4. The
PapiEx homepage is at \n{http://icl.cs.utk.edu/~mucci/papiex/}.

@@

LaMarca's final step was to create a multi-mergesort. This attempted to reduce
the number of cache misses in the final merge steps of tiled mergesort by
merging all the sorted arrays in one step. This required the use of an 8-heap,
which had been developed as part of heapsort.
%TODO this isnt what happened exactly

Once these sorts were written it was attempted to improve them slightly.
Instead of a tiled mergesort, a double-aligned tiled mergesort was written.
This attempted to completely sort in the level 1 cache, then to completely sort
in the level 2 cache. Double-aligned multi-mergesort was a combination of this
and of multi-mergesort. The rewrite was based on the base mergesort, and was
written far more cleanly, being more readable and efficient. These were also
found to reduce the instruction count, as well as the level 2 cache miss count.
These results are discussed in more detail in Section \ref{Mergesort results}.

@@

Radixsort, shellsort and each of the four $O(NlogN)$ sorts were treated in this
fashion.

@@

%TESTING
%TODO a lot of this may be in the first sorts chapter, maybe heapsort.
%discuss: BIG + random.org, 64 bit in lamarca vs 32 bit, do_nothing and how it
%affects cache hits, scaling in the software branch predictors. How to compare
%and contrast varying results (branch vs softwre branhc; P4 vs simplescalar).
%What factors are used for testing: /proc/cpuinfo output, cahce and branch
%predictor data from P4, simplescalar inputs, software predictor, verfication of
%sorts.

@@

As well as the number of cycles, Pentium 4 performance registers can also provide
most of the same information as SimpleScalar, with different parameters. For
example, the cache line of a Pentium 4 machine is 128 bytes whereas 32 byte
cache lines were used in SimpleScalar. The addition of real world data would
strengthen our results.
%TODO not true. we have it
